---
title: "Intermediate GIS and Mapping in R"
subtitle: "Introduction to Leaflet"
date: today
author: "Brittany Barker"
fig-cap-location: top
format:
  html:
    page-layout: full
    code-fold: show
    theme: flatly
    embed-resources: true
execute:
  enabled: true
  echo: false
editor: 
  markdown: 
    wrap: 72
---

```{css}
.figure {
margin: auto;
text-align: center;
}
```

## **Load packages**

```{r, warning = FALSE, message = FALSE, echo=TRUE}
library(here) # Project relative paths
library(dplyr) # Data wrangling
library(sf) # Working with spatial features
library(terra) # Working with rasters
library(ggplot2) # For binning data
library(leaflet) # Creating an interactive map
library(leafem) # Addition Leaflet map features
library(htmltools) # Adding a title to map
```

## **Overview**

In the previous workshop, we used `ggplot2` to create *static* maps of
predicted vs. observed dates of lilac leaf out for 2018. This time, we
will use `leaflet` to make our map *interactive*. For example, we will
add capabilities to zoom, pan, and query the map.

## **Learning objectives**

-   Gain additional practice with spatial operations in R\
-   Create an interactive map in R using the `leaflet` package\
-   Understand situations where an interactive map may be preferable to
    static map and vice versa

## **Introduction**

Leaflet is an [open-source](https://github.com/Leaflet/Leaflet)
JavaScript library for making interactive web maps. After creating a map
widget, one can layer features onto that map widget to improve its
display and interactivity.

The [`leaflet`](https://rstudio.github.io/leaflet/) R package allows you
to create a Leaflet map in R. The package includes powerful and
convenient features for integrating with [Shiny
applications](https://rstudio.github.io/leaflet/articles/shiny.html),
allowing you to create interactive online maps.

## **Make the leaflet map**

### **Creating a basic leaflet map**

Whenever we’re working with `leaflet`, we use the function `leaflet()`
to initialize a map widget. Running this code will simply load a gray
panel void of any geographic features.

```{r, echo=TRUE, warnings=FALSE}
leaflet()
```

Before we define our areas of interest (states) and layer on data, we
can add a "basemap" using the map widget. Basemaps consist of map tiles,
or individual map sections that join together to form a composite
picture. We can add a basemap to the map widget using
[`addTiles()`](https://rstudio.github.io/leaflet/articles/basemaps.html).
Running this without adjustment will result in map of the entire world.

By default, `addTiles()` generates a basemap using
[`OpenStreetMap`](https://www.openstreetmap.org/#map=4/38.01/-95.84)
tiles. There are many other options to choose from. You can use
`addProviderTiles()` instead of `addTiles()` to pick from among a
pre-specified set of third-party options, or you can call a tile set
from a URL by providing a link in `addTiles()`.

```{r, echo=TRUE, warnings=FALSE}
leaflet() %>% 
  addProviderTiles("OpenStreetMap")
```

Our data are for the conterminous U.S. (CONUS). We can use the
[`setMaxBounds()`](https://www.rdocumentation.org/packages/leaflet/versions/2.2.2/topics/setView)
function in combination with the `minZoom` option to restrict the map to
CONUS only. Many more
[options()](https://rstudio.github.io/leaflet/articles/widget.html) for
the map widget are available.

```{r, echo=TRUE, warnings=FALSE}
leaflet(options = leafletOptions(minZoom = 3.5)) %>% 
  setMaxBounds(lng1 = -126, lat1 = 25, lng2 = -66.5, lat2 = 49.5) %>% 
  addProviderTiles("OpenStreetMap")
```

### **Import and format data**

First, import the raster data for model predictions of lilac leaf out
for 2018 (raster data: `lilac_2018.tiff`) using the `rast()` function in
`terra`. Name the raster `lilacModel_r` and round it to the `0` decimal
place because the data are days of the year (no decimals).

```{r, echo=TRUE, warnings=FALSE}
# Raster data: model predictions for lilac leaf out
# Use the 'rast()' function to import the raster
# lilacModel_r 

# Round values down because these are days of the year 

```

Next, import phenometric data for observed lilac leaf out dates for 2018
using `st_read()`. We converted the for formatted data to a shapefile in
the previous exercise. Name the object `lilacObs_sf`.

```{r, echo=TRUE, warnings=FALSE}
# Phenometric data (shapefile produced in previous exercise)

```

Finally, import the state boundary data (`states.shp`) and crop it to
the same extent as the raster using `st_crop()`. Name the object
`states`.

```{r, echo=TRUE, warnings=FALSE}
# US states feature


# Crop to same extent as raster and project 

```

As with our `ggplot` maps, we'll view months instead of days of the
year. The code below creates a data frame to make it easier to define
factor levels for the spatial data.

```{r, echo=TRUE, warnings=FALSE}
# Data frame needed to create pretty plots
factors_df <- data.frame(
  # Day and month of year
  leaf_doy = 1:365, 
  leaf_month = ggplot2::cut_interval(1:365, 12)) %>%
  # Bin dates by month and re-format labels to remove brackets, parentheses, etc.
  mutate(leaf_month = format(
    as.Date(
      leaf_doy - 1, origin = "2018-01-01"), "%b"),
      leaf_month = gsub("\\(|\\]|\\[", "", leaf_month)
    ) %>%
  mutate(leaf_month = gsub(",", "-", leaf_month))

# Convert month to a factor so they're in the right order on plots
# Factor levels are ordered by day of the year
factors_df$leaf_month <- factor(
  factors_df$leaf_month, 
  levels = unique(factors_df$leaf_month[order(factors_df$leaf_doy)]))
```

Next, convert leaf out month data in both datasets to factor format for
plotting. As before, use the `factors_df` data frame to define factor
levels.

```{r, echo=TRUE, warnings=FALSE}
# Define factor levels for the raster
lilacModel_r <- as.factor(lilacModel_r)
levels(lilacModel_r) <- factors_df

# Define factor levels for the observations (sf object)
lilacObs_sf$leaf_month <- factor(
  lilacObs_sf$leaf_month, levels = unique(factors_df$leaf_month))
```

In R, `leaflet` uses several
[`color()`](https://rstudio.github.io/leaflet/articles/colors.html)
functions to map variables to color. First we need to define the palette
and then join that information to the `factors_df` data frame so that
each value (day of year) has a color associated with it.

```{r, echo=TRUE, warnings=FALSE}
# Create a data frame that defines a color for each month of the year
pal_df <- data.frame(
  pal = c("#1f83b4", "#12a2a8", "#2ca030", "#78a641", "#bcbd22", "#ffbf50",
         "#ffaa0e", "#ff7f0e","#d63a3a", "#c7519c", "#ba43b4", "#8a60b0"), 
  leaf_month = levels(factors_df$leaf_month))

# Join this data frame with the `factors_df` data frame
factors_df <- left_join(factors_df, pal_df, by = "leaf_month")
```

Next, we create two separate color palettes to accommodate the different
data formats for predicted vs. observed leaf out dates (raster vs.
vector, respectively). Both palettes are creating using
[`colorFactor()`](https://campus.datacamp.com/courses/interactive-maps-with-leaflet-in-r/plotting-points?ex=15)
because we're working with factor data (months).

```{r, echo=TRUE}
# Colors for raster (predicted leaf out)
palRast <- colorFactor(
  palette = factors_df$pal, 
  domain = factors_df$leaf_doy,
  na.color = "transparent" # Make missing values transparent
)

# Colors for points = sf object (observed leaf out)
palPts <- colorFactor(
  palette = unique(factors_df$pal),
  domain = lilacObs_sf$leaf_month)
```

------------------------------------------------------------------------

### **Add basemap and raster data**

Raster data are displayed on a leaflet map using
[`addRasterImage()`](https://rstudio.github.io/leaflet/reference/addRasterImage.html).
The function accepts both `SpatRaster` and `RasterLayer` (from the old
`raster` package) objects. The `layerId` is a unique identifier for a
specific layer. The `group` allows interaction with other map elements
(added below). The opacity is set to `0.6` to allow visibility of
features in the base map (i.e., underneath the raster).

```{r, echo=TRUE, warnings=FALSE}
leaflet(options = leafletOptions(minZoom = 3.5)) %>% 
  setMaxBounds(lng1 = -126, lat1 = 25, lng2 = -66.5, lat2 = 49.5) %>% 
  # Tiles
  addProviderTiles("OpenStreetMap") %>% 
  # Add raster (model predictions of leaf out)
  addRasterImage(lilacModel_r, 
                 project = TRUE, 
                 colors = palRast,
                 opacity = 0.6,
                 layerId = "Predicted",
                 group = "Predicted")
```

------------------------------------------------------------------------

#### **Practice 1** Explore basemaps

Replace "OpenStreetMap" with three different basemaps. A live visual
index of every variant available is in the [Leaflet Provider Demo
Gallery](https://leaflet-extras.github.io/leaflet-providers/preview/).

```{r}
addProviderTiles("CartoDB.Positron")
addProviderTiles("Esri.WorldImagery")
addProviderTiles("OpenTopoMap")
```

Questions:

**Question 1.** Which basemap would work best for displaying the raster
data? Why?

**Question 2.** How does changing the color and opacity of raster data
affect base map readability?

------------------------------------------------------------------------

### **Control the starting view**

You can set the view to a particular area of interest using
[`setView()`](https://www.rdocumentation.org/packages/leaflet/versions/2.2.2/topics/setView).
Try setting the view for the Pacific Northwest using the argument below.

```{r, echo=TRUE, warnings=FALSE}
# setView(lng = -120, lat = 44, zoom = 6) 
```

------------------------------------------------------------------------

#### **Practice 2** Control the starting view

Center the map on your home state and zoom in.

```{r}

```

Questions:

**Question 3.** What happens when you increase the zoom level?

**Question 4.** What happens when you remove `setMaxBounds()`?

------------------------------------------------------------------------

### **Image query**

The
[`addImageQuery()`](https://rdrr.io/github/r-spatial/leafem/man/addImageQuery.html)
function from the [`leafem`](https://rdrr.io/github/r-spatial/leafem/)
package allows you to query specific pixels. Note, however, that the
tools returns the day of year rather than the month.

```{r, echo=TRUE, warnings=FALSE}
leaflet(options = leafletOptions(minZoom = 3.5)) %>% 
  setMaxBounds(lng1 = -126, lat1 = 25, lng2 = -66.5, lat2 = 49.5) %>% 
  # Tiles
  addProviderTiles("OpenStreetMap") %>% 
  # Add raster (model predictions of leaf out)
  addRasterImage(lilacModel_r, 
                 project = TRUE, 
                 colors = palRast,
                 opacity = 0.6,
                 group = "Predicted",
                 layerId = "Predicted") %>% 
  # Raster query tool
  addImageQuery(lilacModel_r, 
                project = TRUE, 
                group = "Predicted",
                layerId = "Predicted")
```

------------------------------------------------------------------------

#### **Practice 3** Investigate the raster

Use the query tool to identify some of the earliest and latest predicted
leaf-out dates.

```{r}

```

Questions:

**Question 5.** Where are the earliest vs latest predictions located?

**Question 6.** Do you have ideas about how to display calendar date
units?

------------------------------------------------------------------------

### **Add boundary features**

Next, we can add our vector data (`sf` objects), including the states
feature and the points where lilac leaf out was observed. The
`addPolylines()` is used for states (boundaries only) whereas
`addCircleMarkers()` is used for the points. Both functions are part of
the
[`addControl()`](https://www.rdocumentation.org/packages/leaflet/versions/2.2.2/topics/addControl)
collection in `leaflet`. The color palette created above is used to fill
each point accordingly.

```{r, echo=TRUE, warnings=FALSE}
  # # Add state lines
  # addPolylines(data = states, 
  #              opacity = 0.25, 
  #              color = "black", 
  #              weight = 1.25) %>%
  # # Add points (observations of leaf out)
  # addCircleMarkers(
  #   data = lilacObs_sf,
  #   group = "Observed",
  #   popup = ~paste("Leaf-out month:", leaf_month),
  #   color = "black",
  #   fillColor = ~palPts(leaf_month),
  #   fillOpacity = 1,
  #   radius = 3,
  #   weight = 0.5) 
```

------------------------------------------------------------------------

#### **Practice 4** Conditional layers

Let's say you want to show a specific layer at certain zoom levels. For
example, someone may want to see county boundaries when they zoom in to
the state level. This can be accomplished using `groupOptions()` and the
`zoomLevels` argument.

Import the county boundary data (`counties.shp`) and crop it to the same
extent as the raster. Name the object `counties`.

```{r, echo=TRUE, warnings=FALSE}
# US counties feature

# Crop to same extent as raster and project 
```

You can make the counties layer conditional using the following
argument:

```{r, echo=TRUE}
# groupOptions("Counties", zoomLevels = 6:18)
```

Now add the counties layer to your map and use the above conditional
command.

```{r, echo=TRUE}

```

------------------------------------------------------------------------

#### **Practice 5** Customize boundaries and points

Experiment with different state boundary styles and point symbols. Hint:
check out the full range of options for graphic elements and layers in
the reference for
[`map-layers`](https://rstudio.github.io/leaflet/reference/map-layers.html).

```{r}

```

Questions:

**Question 7.** Which settings make boundaries and points easiest to
see? Conversely, When might thick boundaries be distracting?

Try adding a popup displaying the observation month to
`addCircleMarkers`.

```{r, echo=TRUE}
# popup = ~paste("Leaf-out month:", leaf_month)
```

**Question 8.** Zoom into a location with a large difference between
predicted and observed leaf out. Do you think this is observer or model
error?

------------------------------------------------------------------------

#### **Practice 6** Add layer controls

Create a control panel that allows users to turn layers on and off.
Below I added code for the model prediction (raster) layer.

```{r, echo = TRUE}
# addLayersControl(
#   overlayGroups = c("Predicted", "Observed"),
#   options = layersControlOptions(collapsed = FALSE)
# )
```

Questions:

**Question 9.** Can you figure out how to add layer controls for the
other spatial features?

------------------------------------------------------------------------

### **Add a legend**

The map isn't very informative without a legend. The
[`addLegend()`](https://rstudio.github.io/leaflet/reference/addLegend.html)
function `leaflet` can solve this problem. The
[`addLeafLegends`](https://rdrr.io/cran/leaflegend/man/addLeafLegends.html)
package has additional functions for customizing legends. The raster and
point data have the same color scale, so only a single legend is needed.

```{r, echo=TRUE}
# Add legend
# addLegend(
#   data = lilacObs_sf,
#   pal = palPts,
#   values = ~leaf_month,
#   position = "bottomleft",
#   title = "Month",
#   opacity = 0.9
# ) 
```

### **Add a title**

The map looks pretty good now, but what about a title? Below are some
HTML tags, which are building blocks that define the structure and
content of a webpage. The style element (`tags$style`) tell `leaflet`
exactly how the title should be displayed, including its placement on
the map, size, padding, colors, etc. The

<div>

tag (`tags$div`) defines a division or a section in an HTML document.
Here the section is our stylized title.

```{r, echo=TRUE, warnings=FALSE}
# HTML tags for the title style
tag.map.title <- tags$style(HTML("
  .leaflet-control.map-title {
  left:50%;
  transform: translate(-50%);
  position: fixed !important;
  text-align: center;
  padding-left: 10px;
  padding-right: 10px;
  background: rgba(255,255,255,0.75);
  font-weight: bold;
  font-size: 14px;
  color: #48494B;
  }
"))

# HTML tag for the title itself
title <- tags$div(
  tag.map.title, HTML("Predicted vs. Observed Leaf Out in Lilac in 2018")
)
```

You can add the title using the code below.

```{r, echo=TRUE}
# Map title
# addControl(title, 
#            position = "topright", 
#            className = "map-title") 
```

------------------------------------------------------------------------

#### **Practice 8** Wrap-up

There are numerous other great map features in `leaflet`. Try adding the
features below.

**1.** Display the coordinates as you move the cursor over the map using
`addMouseCoordinates()`.

**2.** Add a scale bar using `addScaleBar()`.

**3.** Add a mini map using `addMiniMap()`.

------------------------------------------------------------------------

Add your final leaflet map code below to produce a beautiful map! 

```{r, echo=TRUE, warnings=FALSE}

```

------------------------------------------------------------------------

## **Wrap up**

The "Introduction to GIS and Mapping in R" workshop focused on making
static maps using `ggplot2`. Now that you've completed this exercise,
consider the questions below.

**Question 10.** What information is easier to explore with an
interactive map?

**Question 11.** What information is easier to communicate with a static
map?

**Question 12.** When would you choose `ggplot2` instead of `leaflet`?

**Question 13.** When would you choose `leaflet` instead of `ggplot2`?

## **Potentially useful resources about `leaflet`**

**Book chapters**\
- [Making maps in
R](https://bookdown.org/nicohahn/making_maps_with_r5/docs/leaflet.html)
by N. Hahn\
- [Maps, R Not the Best
Practices](https://bookdown.org/voevodin_nv/R_Not_the_Best_Practices/maps.html)
by N. Voevodin (2020)\

**Tutorials/vignettes**\
- [Introduction to
leaflet](https://rstudio.github.io/leaflet/articles/leaflet.html) by J.
Cheng\
- [Leaflet package in
R](https://www.datacamp.com/courses/interactive-maps-with-leaflet-in-r)
by geeksforgeeks (2024)\
- [Data Scientist as Cartographer: An Introduction to Making Interactive
Maps in R with
Leaflet](https://library.virginia.edu/data/articles/data-scientist-as-cartographer-an-introduction-to-making-interactive-maps-in-r-with-leaflet)
by J. Goldstein-Greenwood (2020)\
- [Interactive maps with leaflet in
R](https://r-charts.com/spatial/interactive-maps-leaflet/) by R Charts
(2025)\
- [Leaflet in
R](https://sesync-ci.github.io/leaflet-in-R-lesson/index.htm) by SESYNC\
- [Interactive maps with leaflet - R for
journalists](https://learn.r-journalism.com/en/mapping/leaflet_maps/leaflet)
by J. Tran (2018)\

## **Acknowledgements**

This work was supported by the USDA National Institute of Food and
Agriculture (NIFA) Agriculture and Food Research Initiative program
(grant no. 2022-68013-37138) and USDA NIFA Crop and Pest Management
(CPPM) program (grant no. 2024-70006-43503).

</div>
