---
title: "Intermediate GIS and Mapping in R"
subtitle: "Raster Data Analysis"
date: today
author: "Brittany Barker"
fig-cap-location: top
format:
  html:
    page-layout: full
    code-fold: show
    theme: flatly
    embed-resources: true
execute:
  enabled: true
  echo: false
editor: 
  markdown: 
    wrap: 72
---

```{css}
.figure {
margin: auto;
text-align: center;
}
```

## **Load packages**

```{r, warning = FALSE, message = FALSE, echo=TRUE}
library(here) # Project relative paths
library(dplyr) # Data wrangling
library(sf) # Working with spatial features
library(terra) # Working with rasters
library(ggplot2) # Make a plot
```

## **Overview**

So far, we have focused on mapping vector and raster data. This exercise
will focus on conducting raster analyses using built-in and custom
functions.

## **Learning objectives**

-   Apply functions to rasters\
-   Write custom functions for raster analyses\
-   Perform raster time-series analyses\
-   Extract and summarize raster values by polygons\
-   Visualize spatial and temporal trends

## **Introduction**

The `terra` package is powerful and efficient at analyzing raster and
vector spatial data. It provides tools for importing, manipulating,
visualizing, and exporting large geospatial datasets while optimizing
memory usage and computational performance. For raster analysis, `terra`
supports common operations such as extraction, cropping, masking,
resampling, reclassification, focal (neighborhood) analyses, terrain
calculations, and raster algebra. This exercise will focus on using
built-in `terra` functions as well as custom function for analyzing
raster data.

## **Exercise 1: Extract and analyze raster data**

Here, we'll use the `extract` function to extract model predictions
(predicted day of year) from the precise location where lilac
observations were collected (observed day of year). This allows us to
further compare the two datasets.

Use code from previous exercises to import the `lilac_2018_model` raster
(model predictions of leaf out) and the phenometric data for lilac leaf
out (observed dates of leaf out).

```{r, echo=TRUE, warnings=FALSE}
# Use the 'rast()' function to import the raster
lilacModel_r <- rast(here("data", "lilac_2018_model.tiff"))

# Phenometric data (shapefile produced in previous exercise)
lilacObs_sf <- st_read(here("data", "lilac_2018_obs.shp"))
```

The `extract()` function in `terra` extracts predictions from location
data, including `sf` point files. The first argument is the raster, the
second the locations.

```{r, echo=TRUE, warnings=FALSE}
# Extract model predictions for each observation location
lilacObs_extr <- extract(lilacModel_r, lilacObs_sf)  
```

Sometimes it's easier to work with the results in a data frame, which
we'll do below. By default, `terra` names the extracted data as `layer`,
so we'll make the column name more informative (`leaf_doy`).

```{r, echo=TRUE}
lilacObs_extr <- lilacObs_extr %>%
  as.data.frame() %>%
  rename("leaf_doy" = "layer")
```

Next, we can estimate the extent to which the model under-predicts vs.
over-predicts the date of lilac leaf out by subtracting the `leaf_doy`
columns.

```{r, echo=TRUE}
# Differences (in days) between predicted and observed lilac leaf out
differences <- lilacObs_extr$leaf_doy- lilacObs_sf$leaf_doy
```

Create a histogram of differences using the `hist()` function in base
`R.` It might be helpful to define a sequence of breaks.

```{r, echo=TRUE, warnings=FALSE}
hist(differences, breaks = seq(-250, 50, 25))
```

Now calculate some simple summary statistics of the differences.

```{r, echo=TRUE, warnings=FALSE}
# Mean
mean(differences)
# Range
range(differences)
# Median
median(differences)
```

------------------------------------------------------------------------

**Question 1** What information do your simple analyses provide that
can't be seen on a map? Overall, what do the results say about the
performance of the model for lilac leaf out?

Answer: On average, predicted dates were 13.71 days earlier (range =
-237 to 48 days) than observed dates, which suggests that the model is
under-predicting the date of lilac leaf out. However, the extreme
outliers are likely due to observer error. The model actually seems to
perform pretty well if we reduce the impact of outliers by calculating
the median difference, which is only 1 day.

------------------------------------------------------------------------

## **Exercise 2: Apply custom functions to raster data**

Custom functions allow us to:

-   Avoid repeating code
-   Improve readability
-   Increase reproducibility
-   Apply the same workflow to many rasters
-   Simplify complex analyses

## **(1) Evaluating temperature anomalies**

We'll move on from the lilac modeling data set to climate change.
Specifically, well ask this question: for each year between 2020 and
2024, which OR counties experienced the largest July temperature
anomalies relative to the 30-year average?

### **Data preparation**

We will use three data sets:

1)  Tmax for July for 2020, 2021, 2022, 2023 and 2024 (5 rasters)
2)  Average Tmax for July between 1981-2010 (1 raster)
3)  U.S. county simple features (sf) object

Below a list of raster files for July in individual years is created,
plus the 30-year average file.

```{r}
# List of files for July
Jul_fls <- list.files(
  here("data", "prism_data", "july"),
  pattern = glob2rx("*202*.tif$*"),
  full.names = TRUE
)

# 30-year normals
Jul_30yrAv_fl <- here("data", "prism_data", "normals", "PRISM_tmax_30yr_normal_4kmM2_07_bil.bil")
```

Take a peek at the raster data - there are 872,505 raster cell values

```{r, echo=TRUE}
Jul_30yrAv <- rast(Jul_30yrAv_fl)

plot(Jul_30yrAv)
```

Import the county boundary data (`counties.shp`) and filter out counties
for OR.

```{r, echo=TRUE, warnings=FALSE}
# US counties feature
counties <- st_read(here("data", "counties.shp"))

# Subset (filter) Oregon counties only
countiesOR <- counties %>% 
  filter(stat_nm == "Oregon")
```

### **Function: temperature anomaly by county**

The `top_county_func` function determines which Oregon counties had the
largest difference in July temperature compared to the 30 year average.
The `n` argument defines the top number of counties with the largest
anomaly to return.

```{r}
top_county_func <- function(i, n = 5) {

  # Import rasters
  r <- rast(Jul_fls[i])
  r_avg <- rast(Jul_30yrAv_fl)
  
  # Calculate the temperature anomaly in degrees Fahrenheit 
  anom <- (r * 1.8 + 32) - (r_avg * 1.8 + 32)
  names(anom) <- "value"
  
  # Calculate the anomaly in OR counties
  county_vals <- extract(anom, countiesOR, fun = mean)
  #county_vals <- extract(anom, countiesOR, fun = median)
  
  # Sort the results and extract the counties with the largest anomalies
   anom_cnty <- county_vals %>%
    mutate(
      county = countiesOR$name,
      anomaly_F = value
    ) %>%
    arrange(desc(abs(anomaly_F))) %>%
    slice(1:n) %>%
    select(county, anomaly_F)
   
   return(anom_cnty)
}
```

### **Apply the function across years**

Use `lapply` to apply the function across `1:length(July_fls)`.

```{r, echo=TRUE, warnings=FALSE}
results <- lapply(1:length(Jul_fls), top_county_func)
results
```

------------------------------------------------------------------------

**Question 1.** Which OR county had the largest difference (anomaly) in
temperature (F) from the 30-year average for July?

The output shows which 5 OR counties had the largest difference
(anomaly) in temperature (F) from the 30 year average for July. The
largest anomalies are for `2021`, a year with record-breaking heat for
Oregon.

#### **Practice 1** Modify the raster function

**1.** Add median, standard deviation, and maximum anomaly values.

**2.** Return counties with the most negative anomalies.

**3.** Modify the function so it returns the top 10 counties instead of
5.

------------------------------------------------------------------------

## **(2) Direction of change in annual Tmax**

How has annual maximum temperature (Tmax) changed across Oregon between
1990 and 2019?

Load annual data.

```{r, echo=TRUE}
USA_fls <- list.files(
  here("data", "prism_data", "annual"),
  pattern = ".bil$",
  full.names = TRUE
)
```

### **Function: regression model of Tmax**

The `slope_func` function below loads, crops, and subsets annual Tmax
rasters and then runs a regression model for each raster grid cell. The
results will provide insight into the direction of change in annual Tmax
for each decade.

```{r, echo=TRUE}
slope_func <- function(years) {
  
  # Get rasters and crop to OR, and retain layers of corresponding years
  USA_r <- rast(USA_fls)
  
  # Crop the raster data
  OR_r <- crop(
    USA_r,
    ext(-124.7294, -116.2949, 41.7150, 46.4612)
  )
  
  # Subset raster layer for the year range
  OR_r <- OR_r[[years]]
  
  # Run a regression model for each cell of the raster stack
  rm <- regress(
    OR_r,
    1:nlyr(OR_r),
    na.rm = TRUE
  )
   return(rm)
}
```

Analyze the trend of annual Tmax for each decade.

```{r, echo=TRUE}
# Decades
year_sets <- list(
  1:10, # 1990-1999
  11:20, # 2000-2010
  21:30 # 2011-2020
)
```

Apply the function below.

```{r, echo=TRUE}
slopes <- lapply(year_sets, slope_func)
```

Take a look at the trend results. What does each list element correspond
to?\

```{r, echo=TRUE}
slopes
```

For increased clarity, we can rename the list elements. These can be
used for titles for creating basic plots in `terra`.

```{r, echo=TRUE}
names(slopes) <- c( "OR_1990_1999",  "OR_2000_2010",  "OR_2011_2020")
```

Let's look at the results. Annual Tmax exhibited different trends
throughout OR if we look at each decade separately, but temperatures are
generally increasing in many areas. The `x` layer in each raster
corresponds to the slope - i.e., the trend in annual Tmax for the
respective time periods. We can plot these results by referencing the
second layer for each list element.

```{r}
plot(c(slopes[[1]][[2]], slopes[[2]][[2]], slopes[[3]][[2]]),
     plg=list(title="Slope"), 
     main = names(slopes))
```

**Question 2.** Overall, which decade did Tmax increase at the fastest
rate?

------------------------------------------------------------------------

#### **Practice 2** Modify the trend function

**4.** Change the extent to a different state.

**5.** Change the time frame to a longer period, such as `20` or `30`
years.

------------------------------------------------------------------------

Visualize the Tmax trends by Oregon county for 2011-2020. Here we
convert `countiesOR` to a `SpatVector` format, which is designed for
vector data in the `terra` package. You can save the results as a
polygon and easily plot the results. By default, the output column name
is `x`.

```{r, echo=TRUE}
# Convert the sf feature to a SpatVector
countiesOR_v <- vect(countiesOR)

# Calculate zonal stats
slope_zonal <- zonal(slopes[[3]][[2]], countiesOR_v, 
                     fun = "mean", as.polygons = TRUE)

# Plot
plot(slope_zonal, y = "x", 
     plg=list(title="Zonal mean"))
```

------------------------------------------------------------------------

#### **Practice 3** Make a nicer plot

**6.** Practice your `ggplot2` mapping skills to make a nicer map!
(Hint: the data is `countiesOR_z` and values are `mean`). A nice
addition could be a title like "Mean Rate of Tmax Change in Oregon
(2011-2020)", and you can visualized county names using
`geom_sf_text()`.

```{r, echo = TRUE}
# Bind the mean value column to the OR counties feature
countiesOR_z <- bind_cols(countiesOR, mean = slope_zonal$x)

# Make a ggplot
ggplot(data = countiesOR_z) +
  geom_sf(aes(fill = mean), color = "white", size = 0.2) +
  geom_sf_text(aes(label = name), size = 1.5) +
  scale_fill_viridis_c(name = "Zonal Mean") + # Clean, continuous color palette
  theme_minimal() +
  labs(
    title = "Mean Rate of Tmax Change in Oregon (2011-2020)"
  )
```

**Question 3.** Which county experienced the largest increase in Tmax
between 2011-2020?

## **Wrap up**

We focused on using `terra` to extract and analyze raster and vector
data. The package is large and extensive, with dozens (or hundreds!) of
functions, and an ability to do parallel processing to speed up
operations. Thus, it's very powerful and an integral to working with
rasters in R.

## **Potentially useful tutorials, free books, etc.**

**Compilation of various resources**\
- [R Spatial data science blogs](https://r-spatial.org/)

**Books**\
- [Geocomputation with R](https://r.geocompx.org/) by R. Lovelace et al.
(2025)\
- [Geographic Data Science with
R](https://bookdown.org/mcwimberly/gdswr-book/) by M.C. Wimberly (2023)\

**Tutorials/vignettes**\
- [Spatial Data Science with R and "terra"](https://rspatial.org/) by R.
Hijmans (2019-2025)\
- [Geospatial workskop](https://datacarpentry.org/lessons/#geospatial)
by NEON\
- [Introduction to Spatial Raster Data Using the terra
Package](https://rpubs.com/GeospatialEcologist/SpatialRaster) by J.
Carrell (2023)\
- [Introduction to rasters with
terra](https://emlab-ucsb.github.io/intro-terra/) by J. Flower\
- [Intro to GIS in R](https://rpubs.com/hannah0/GIS_with_terra) by H.
Owens\
- [Raster analysis with
terra](https://www.wvview.org/os_sa/15_Raster_Analysis_terra.html) by
WVview

## **Acknowledgements**

Development of this workshop was supported by the USDA National
Institute of Food and Agriculture (NIFA) Agriculture and Food Research
Initiative program (#2022-68013-37138) and USDA NIFA Crop and Pest
Management program (#2024-70006-43503).
